---
title: Foundations of Probability for Finite Sample Spaces
author: Mathematics for Computation and Data Science
date: Arnob Roy -- Last Revised on June 24, 2021
output: pdf_document
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{flushleft}\LARGE \scshape}
  - \posttitle{\end{flushleft}}
  
  - \preauthor{\begin{flushleft}\large}
  - \postauthor{\end{flushleft}}
  
  - \predate{\begin{flushleft}\large}
  - \postdate{\end{flushleft}}
---

# R Script Problems

## Problem 1.

```{r}
Q1 <- c(100,0,0,0)
Q2 <- c(100,0,0,0)

scores <- expand.grid(Q1, Q2)

scores <- data.frame(scores, rowSums(scores), rowSums(scores)/2, (scores$Var2-scores$Var1)/100+1)
names(scores) <- c("Q1", "Q2", "Sum", "X", "Y")

scores <- data.frame(scores, scores$X*scores$Y)
names(scores) <- c("Q1", "Q2", "Sum", "X", "Y", "XY")

expected.X <- sum(scores$X/16)
expected.Y <- sum(scores$Y/16)
expected.XY <- sum(scores$XY/16)

expected.XY == expected.X*expected.Y #Uncorrelated

scores <- data.frame(scores, scores$X*scores$X, scores$Y*scores$Y)
names(scores) <- c("Q1", "Q2", "Sum", "X", "Y", "XY", "Xsquared", "Ysquared")
scores <- data.frame(scores, scores$Xsquared*scores$Ysquared)
names(scores) <- c("Q1", "Q2", "Sum", "X", "Y", "XY", "Xsquared", "Ysquared", "XY squared")

expected.Xsq <- sum(scores$Xsquared/16)
expected.Ysq <- sum(scores$Ysquared/16)
expected.XY.sq <- sum(scores$`XY squared`/16)

expected.XY.sq == expected.Xsq*expected.Ysq #Not Independent

scores
```

Since $\mathbb{E}(XY) = \mathbb{E}(X)\mathbb{E}(Y)\to$ $X$ and $Y$ are uncorrelated.

However, $\mathbb{E}(X^2Y^2) \ne \mathbb{E}(X^2)\mathbb{E}(Y^2)\to$ $X$ and $Y$ are not independent.



## Problem 3. Permutation Test

After loading the data set containing the birth weights of babies born in North Carolina, we find that the average male baby is 103.2632 ounces heavier than a female baby.

```{r}
#Load the Data
NC <- read.csv("NCBirths2004.csv"); head(NC)

#Average Birth Weight of a Male
idx <- which(NC$Gender == "Male")
mean(NC$Weight[idx])

#Average Birth Weight of a Female
idx.female <- which(NC$Gender == "Female")
mean(NC$Weight[idx.female])

#Difference in birth weight
mean(NC$Weight[idx])-mean(NC$Weight[idx.female])
```

**Conducting a Permutation Test** (Inspired by Script 2D, and Chihara and Hesterberg)

```{r}
#Record our observed difference in the weights
observed.diff <- mean(NC$Weight[idx])-mean(NC$Weight[idx.female])

#Random Sampling
sum(NC$Gender == "Male") #488 M
sum(NC$Gender == "Female") #521 F
Gender <- sample(NC$Gender); head(Gender)
sum(Gender == "Female")
sum(Gender == "Male")
FemaleAvg <- sum(NC$Weight*(Gender=="Female"))/sum(Gender=="Female"); FemaleAvg
MaleAvg <- sum(NC$Weight*(Gender=="Male"))/sum(Gender=="Male"); MaleAvg
MaleAvg - FemaleAvg

#Let's permute this 10^4 times
N <- 10000
permuted.diff <- numeric(N)

for(i in 1:N){
  Gender <- sample(NC$Gender); Gender
  FemaleAvg <- sum(NC$Weight*(Gender=="Female"))/sum(Gender=="Female")
  MaleAvg <- sum(NC$Weight*(Gender=="Male"))/sum(Gender=="Male")
  permuted.diff[i] <- MaleAvg - FemaleAvg
}
```

```{r}
#Average of the Permutation Distribution
mean(permuted.diff)

#Histogram of the Permutation Distribution
hist(permuted.diff, breaks = "FD")
abline(v = observed.diff, col = "red")

#What is the probability that this difference arose from a random subset?
p <- (sum(permuted.diff >= observed.diff)+1)/(N+1); p
```

\newpage

## Problem 4. Permutation Test comparing the Difference between Celtiberian Champions and Celtiverian Swordsmen

```{r}
#Load the Data
Battles <- read.csv("Battle.csv"); head(Battles)

#Average Kills of MEZU/BRIG
MEZU <- which(Battles$Abbr == "MEZU"); head(MEZU)
BRIG <- which(Battles$Abbr == "BRIG"); head(BRIG)
mean(Battles$Kills[MEZU])
mean(Battles$Lost[MEZU])
mean(Battles$Kills[BRIG])
mean(Battles$Lost[BRIG])

#Difference Kills - Lost for each unit
observed.mezu <- mean(Battles$Kills[MEZU]) - mean(Battles$Lost[MEZU]); observed.mezu 
observed.brig <- mean(Battles$Kills[BRIG]) - mean(Battles$Lost[BRIG]); observed.brig
```

```{r}
#Random Sampling
Unit <- sample(Battles$Abbr)
KillsAvg.MEZU <- sum(Battles$Kills*(Unit == "MEZU"))/sum(Unit == "MEZU"); KillsAvg.MEZU
LostAvg.MEZU <- sum(Battles$Lost*(Unit == "MEZU"))/sum(Unit == "MEZU"); LostAvg.MEZU
perform.MEZU <- KillsAvg.MEZU - LostAvg.MEZU; perform.MEZU

#Let's Conduct a Permutation Test for our MEZU unit, based on Kills-Lost, 10^5 times
N <- 10^5
diffs.MEZU <- numeric(N)
for (i in 1:N){
  Unit <- sample(Battles$Abbr)
  KillsAvg.MEZU <- sum(Battles$Kills*(Unit == "MEZU"))/sum(Unit == "MEZU")
  LostAvg.MEZU <- sum(Battles$Lost*(Unit == "MEZU"))/sum(Unit == "MEZU")
  perform.MEZU[i] <- KillsAvg.MEZU - LostAvg.MEZU; perform.MEZU
}

Unit <- sample(Battles$Abbr)
KillsAvg.BRIG <- sum(Battles$Kills*(Unit == "BRIG"))/sum(Unit == "BRIG"); KillsAvg.BRIG
LostAvg.BRIG <- sum(Battles$Lost*(Unit == "BRIG"))/sum(Unit == "BRIG"); LostAvg.BRIG
perform.BRIG <- KillsAvg.BRIG - LostAvg.BRIG; perform.BRIG

#Let's Conduct a Permutation Test for our MEZU unit, based on Kills-Lost, 10^5 times
N <- 10^5
diffs.BRIG <- numeric(N)
for (i in 1:N){
  Unit <- sample(Battles$Abbr)
  KillsAvg.BRIG <- sum(Battles$Kills*(Unit == "BRIG"))/sum(Unit == "BRIG")
  LostAvg.BRIG <- sum(Battles$Lost*(Unit == "BRIG"))/sum(Unit == "BRIG")
  perform.BRIG[i] <- KillsAvg.BRIG - LostAvg.BRIG
}
```

```{r}
mean(perform.MEZU)
hist(perform.MEZU, breaks = "FD")
abline(v = observed.mezu, col = "red")
p.mezu <- (sum(perform.MEZU >= observed.mezu)+1)/N+1; p.mezu

mean(perform.BRIG)
hist(perform.BRIG, breaks = "FD")
abline(v = observed.brig, col = "red")
p.brig <- (sum(perform.BRIG >= observed.brig)+1)/N+1; p.brig
```




