---
title: "Math 23c - Week 2"
author: "Paul Bamberg"
output:
  slidy_presentation:
    font_adjustment: +1
    incremental: true
    widescreen: true
  classoption: aspectratio=169
runtime: shiny
---
<style type="text/css">
div.slide p {
  color: DarkBlue;
  font-size:smaller;
}
div.slide h1 {
  color: DarkRed;
  font-size: 100%;
}
div.slide h2 {
  color: DarkGreen;
  font-size: smaller
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The pigeonhole principle for finite sets

* Let $A$ be the set of your pigeons, and let $B$ be the set of pigeonholes in which they live.

* Let $f: A \rightarrow B$ be the function that assigns to each pigeon the pigeonhole in which it lives.

* The pigeonhole principle states that if $|A| > |B|$, then there exists a pigeonhole in which more than one pigeon lives.

* Here is a diagram with $|A| = 4, |B| = 3$ to illustrate this idea.

```{r, echo = FALSE, fig.width = 8, fig.height = 4}
plot(NULL, xlim = c(-.5, 1.5), ylim = c(-.1, 0.8),mar = c(0,0,0,0), xlab = "", ylab = "", xaxt = "n" ,yaxt = "n")
arrows(c(0,0,0),c(0.2, 0.4, 0.6),c(1,1,1),c(0.2, 0.4, 0.6))
arrows(0,0,1,0.2, lty = 3)
points(c(0,0,0,0,1,1,1),c(0, 0.2, 0.4, 0.6, 0.2, 0.4, 0.6))
text(0, 0.7, "A")
text(1, 0.7, "B")
text(0.7, 0, "This pigeon must move into an occupied pigeonhole")
```

* A more formal statement of the pigeonhole principle for the case of finite sets:  
Let $A$ and $B$ be finite sets, with $|A| > |B|.$ 
Then no function $f: A \rightarrow B$ can be injective.

# Role reversal

* We can reverse roles. Here is a diagram with $|A| = 3, |B| = 4$ where now $A$ is the set of pigeonholes, $B$ the set of pigeons.
```{r, echo = FALSE, fig.width = 8, fig.height = 4}
plot(NULL, xlim = c(-.5, 2), ylim = c(-.1, 0.8),mar = c(0,0,0,0), xlab = "", ylab = "", xaxt = "n" ,yaxt = "n")
arrows(c(0,0,0),c(0, 0.2, 0.4),c(1,1,1),c(0.2, 0.4, 0.6))
points(c(0,0,0,1,1,1,1),c(0, 0.2, 0.4, 0, 0.2, 0.4, 0.6))
text(0, 0.7, "A")
text(1, 0.7, "B")
text(1.4, 0, "This pigeonhole is unoccupied")
```

* A more formal statement of this case:  
Let $A$ and $B$ be finite sets, with $|A| < |B|.$   Then no function $f: A \rightarrow B$ can be surjective.
 
# A useful corollary:

* Let $A$ and $B$ be finite sets, with $|A| = |B|.$ Then if $f: A \rightarrow B$ is not surjective, it is not injective.  
We can prove this corollary, using the pigeonhole principle.

* Proof:
  + Let $C\subset B$ be the image of function $f: A \rightarrow B.$
  
  + Since $f$ is not surjective, $|C| < |B|.$
  
  + Given that $|B| = |A|$, $|C| < |A|.$
  
  + In other words $|A| > |C|$ and by the pigeonhole principle $f$ is not injective.

* Contrapositive: Suppose $|A| = |B|$.  
If $f: A \rightarrow B$ is injective, it is also surjective.





# The pigeonhole principle is useful for proving that a "trial and error'' algorithm will surely succeed.

* Give me a set of four positive integers, for example 7, 17, 5, and 11. By adding and subtracting some or all of them, I guarantee to make an integer that is divisible by 15.

* The pigeonholes are the nonzero remainders when an integer is divided by 15. There are 14 of them, 1, 2, ... 14.

* The pigeons are the numbers that result from adding up one or more of the four integers, dividing by 15, and keeping the remainder. 
 
* From $\{7,17,5,11\}$ you can make $2^4 - 1 = 15$ nonempty subsets.  Therefore there are more pigeons than pigeonholes and there must be either zero remainder or a duplicate nonzero remainder. 

# Using R to find the solution

```{r echo = TRUE}
theSet <- c(7, 17, 5, 11)
#Generate a data frame with all ways of including or excluding a number
df <- expand.grid(c(0,1),c(0,1),c(0,1),c(0,1)) #all possible subsets
colnames(df) <- theSet
df$sum <- df[[1]]*theSet[1] + df[[2]]*theSet[2] + df[[3]]*theSet[3] + df[[4]]*theSet[4]
df$remainder <- df$sum %%15  #remainder when subset sum is divided by 15
df[2:16,]   #ignore the empty set 
```

* In this case one duplicate remainder is 5.
It results from the subset $\{5\}$ and from $\{7, 17, 11\}$
So $7 + 17 +11 - 5 = 30$, which is divisible by 15

The other duplicate remainders, 7 and 3, both lead to $17+5 - 7 = 15$

# Finding a good rational approximation to $\pi$

* Find a rational number $q$ for which $|\pi - q|$ is less than 1/10.  
  + The pigeons are the digits after the decimal point in the first 11  multiples of $\pi$
  + The pigeonholes are the 10 digits 0, 1, ... 9.
  + There must be a duplicate digit (in fact, there are three).
  + We can use any pair of duplicates to find a fraction that is close to $\pi$.
$$
\begin{array}{|c|c|c|c|c|c|c|c|c|c|c|}
\hline
0&1&2&3&4&5&6&7&8&9&10\\
\hline
0.00& 3.14 & 6.28 &9.42 &12.56 &15.70 &18.84  & 21.99 & 25.13 & 28.27 & 31.41\\
\hline
0&1&2&4&5&7&8&9&1&2&4\\
\hline
\end{array}
$$
* 2 and 9 lead to a duplicate digit 2 after the decimal point.  
So $9 \pi - 2 \pi = 28.27 -6.28 = 21.99$  
Thus $7 \pi = 21.99$ and $\pi \approx 22/7.$

* The other duplicate pigeons, 1 from $\pi$ and $8 \pi$ and 4 from $3\pi$ and $10 \pi$, both also lead to 22/7




# The generalized pigeonhole principle states  that if $|A| > k|B|$, then there exists a pigeonhole in which more than $k$ pigeons live.

* Proof by contraposition:
  + Let there be $|B| = n$ pigeonholes.
Assume that each pigeonhole $B_i$ has $\leq k$ pigeons.
  + Then $$|A| = \sum_{i=1}^n B_i \leq kn$$
  + Thus $|A| \leq k|B|$, which contradicts $|A| > k|B|$.

* Trivial application. Noah is watching 10 species board the ark. He counts a total of 21 animals. Show that at least one species has exceeded its quota of 2.  
$|A| = 21, |B| = 10$ and so $|A| > 2 |B|$   
It follows that there must be a species with more than 2 representatives.

# A Tricky Application:

* The Massachusetts legislature has passed new blue laws which forbid the teaching of mathematics on three specific days of the week.  
Prove that no matter which three days they choose, there exists a sequence of three consecutive days on at least two of which the teaching of mathematics is forbidden.

  + The pigeonholes are the 7 different 3-day sequences.

  + The pigeons are the mandatory days of rest in these sequences. Note that if math cannot be taught on Tuesday, that leads to a day of rest in three pigeonholes, SMT, MTW, and TWTh.

  + So each day of rest places a pigeon in each of three different pigeonholes.

  + The three days of rest place a total of 9 pigeons into the 7 pigeonholes.

  + It follows that there must be a pigeonhole occupied by more than one pigeon.

  + In other words, there must a sequence of three consecutive days that includes at least two on which the teaching of mathematics is forbidden.

* This approach is a little-known alternative to the celebrated "probabilistic method" invented by Paul Erdös.



# Probability for finite sample spaces -- the simplest case

* The sample space $S$ consists of possible outcomes of an experiment,  
e.g. roll two dice, 1 red, 1 green. 

* An outcome of an "experiment" is denoted by  $\omega$, e.g. "3 on red, 4 on green."

* An "event'' $A$ is a subset of the sample space, e.g. "sum of the two numbers is 7."

* The "sigma field" of events $\mathcal{F}$ to which probabilities can be assigned is assumed to be the "power set" $2^S$: the collection of all possible subsets of $S$.   
So $\mathcal{F}$ includes not only "total is 6" but also "6 the hard way, 3+3." 

* For a simple probability function, assume that each of the $n$ elements in $S$ has probability $1/n$. Then the probability of event $A$ can be determined by counting the number of elements in event (subset) $A$ and dividing by $n$.

* A random variable is a real-valued function on the sample space, e.g. the number of sixes. 


# Rules for a sigma-field in the finite case

* Sigma-field $\mathcal{F}$ is a collection of events: subsets of the sample space $S$. 

* We assign a probability to an event $A$ only if it is in the sigma-field.

* If events $A$ and $B$ are in the  sigma-field $\mathcal{F}$, so is their union  $A \cup B$.

* If events $A$ is in the  sigma-field $\mathcal{F}$, so is its complement $A^c$.


# Sigma-fields for dice

* If we roll two fair dice, one red and one green, an element of the sample space is of the form "2 on red, 5 on green." 

* There are 36 elements in the sample space, each with probability $\frac{1}{36}$

* The sum of the numbers on the dice is a random variable $T$.

* Often only events like $T = 3$ "the total roll is 3" are of interest. 

* For $T=3$ there are two outcomes, 2+1 and 1+3, so the probability is 2/36 = 1/18.

* For $T=7$ there are six outcomes, $6+1, 5+2 \cdots  1+6$, so the probability is 6/36 = 1/6.

* Other events in the sigma-field include "same number on both dice" and "total roll is a prime number."

# The sigma-field and probabilities for Craps

* For the opening roll in Craps, significant events are   
$W$: 7 or 11, shooter wins and  
$L$: 2, 3, or 12, shooter loses.

* There is a third event $P$: 4, 5, 6, 8, 9, or 10. The number becomes the shooter's point and they keep rolling until they rool the point again and win or roll a 7 and lose.

* The events $W, L$, and $P$ form a **partition** of the sample space.

* The probabilities are $\mathbf{P}(W) = 6/36 + 2/26 = 2/9$  
$\mathbf{P}(L) = 1/36 + 2/36 + 1/36 = 1/9$  
$\mathbf{P}(P) = 1 - 2/9 = 1/9 = 2/3$

* There are eight events in the sigma-field.
$$S, \emptyset, W, W^c, L, L^c, W \cup L, (W \cup L)^c$$



# Binomial coefficients as an aid in counting

* If you carry out an experiment where each of a finite number of possible outcomes is equally likely, the probability of a successful outcome is $$\mathbf{P}(\text{Success event})=\frac{\text{Number of successful outcomes}}{\text{Number of possible outcomes}}$$

* Suppose that set $A$ has $m$ elements and set $B$ has $n$ elements.  The Cartesian product $A \times B$ is the set of ordered pairs $(a, b)$, where $a \in A$ and $b \in B$.  
The set $A \times B$ has $mn$ elements.

* If $A$ has $n$ elements and you select a sequence of $r$ elements **with replacement** from $A$, there are $n^r$ possible sequences. 

* If set $A$ has $n$ elements, then the number of distinct sequences of $r$ different elements (chosen without replacement, so no repetition allowed) is
$$n(n-1)(n-2)...(n-r+1)=\frac{n!}{(n-r)!}.$$ 

* If set $A$ has $n$ elements and you select a set of $r$ different elements from $A$, there are $r!$ distinct sequences for each set. Divide the number of sequences by $r!$ to get
$$\binom{n}{r}=\frac{n!}{(n-r)! r!} \text{, the binomial coefficient "n choose r"}.$$



# Sigma-fields and probabilities for coin flips

* Instead of "flipping $n$ coins,'' where it is hard to keep track of which coin is \#1, it is usually easier to think of flipping a single coin $n$ times in succession. Then you get a sequence of results, like HHTHTTT.

* The number of heads is a random variable $X$ on the sample space, and often the only events of concern are of the form $X = k$.
 
* Example: for $n= 6$, find the probability of the event $X = 3$.
 
* Solution: each sequence like HTTHHT has probability $\frac{1}{2^6}$. There are
$$\binom{6}{3} = \frac{6!}{3!} = \frac{6\cdot 5 \cdot 4}{3 \cdot 3 \cdot 2 \cdot 1} = 20$$ such sequences.
So the probability is $\frac{20}{64} = \frac{5}{16}$

* For $n=6$, here are other types of events that must be in the sample space.  
$X \neq 3$, $X$ is odd, $X \leq 3$, $X >3$)

# Generating the sigma-field
This idea will generalize to countably infinite sample spaces (week 3), and even to uncountably infinite sample spaces (week 5).

Once you know the probability $\mathbf{P}( X \leq k)$ for all $k$ , you can use it to find the probability of any other event, like $X >4$ or $3 < X < 6$.

* The event $X > 4$ is the complement of $X \leq 4$. So it is in the sigma-field, and 
$$\mathbf{P}(X > 4) = 1 - \mathbf{P}(X \leq 4)$$

* $$\mathbf{P}(3 < X < 6) = \mathbf{P}(X \leq 5) - \mathbf{P}(X \leq 2)$$
 
* We say that the events $\mathbf{P}( X \leq k)$ "generate the sigma-field".



# Sigma-fields and probabilities for cards

* The usual assumption is that the deck of cards has been shuffled so well that each of the 52! permutations of the cards is equally likely. This requires a minimum of seven shuffles.

* In bridge, a player is dealt a hand of 13 cards. The number of different **sequences** of 13 cards is 
$52 \cdot 51 \cdot 50 \cdots 40 = \frac{52!}{{39!}}$

* Players usually rearrange the cards in their hand, losing all information about the order in which the cards were dealt to them. The number of different **sets** of 13 cards is
$\binom{52}{13} = \frac{52!}{39!13!}$

* Bridge players are especially concerned with the number of cards in each individual suit.
The number of ways to select 5 spades, 4 hearts, 3 diamonds, and 1 club to make a bridge player's hand is
$\binom{13}{5}\binom{13}{4}\binom{13}{3}\binom{13}{1}.$

* Books about bridge often include probability tables for events like "a 5-4-3-1 hand," ignoring which suit has 5 cards, which has 4, etc. There are $4! = 24$ ways to assign suits to 5, 3, 4, and 1; so we multiply the preceding result by 24. But beware of overcounting: for a 5-4-2-2 hand there are only $4\cdot 3 = 12$ ways to assign suits, and for 4-3-3-3 or 7-2-2-2 there are only 4 ways to choose the long suit.


# Technique for calculating with binomial coefficients

* First, cancel complete binomial coefficients.

* Second expand in terms of factorials and clear complex fractions.

* Third, cancel factorials.

* Fourth, cancel factorials that are close together.
 
* Last, expand any remaining factorials as products.



# Sample problem: By what factor is a 4-4-3-2 hand more probable than a 4-3-3-3 hand?
* There are 12 ways to choose suits for 4-4-3-2, 4 ways to choose suits for 4-3-3-3.

* Start by canceling the binomial coefficients that occur in numerator and denominator:
$$\frac{12 \binom{13}{4} \binom{13}{4} \binom{13}{3} \binom{13}{2}}{4\binom{13}{4} \binom{13}{3} \binom{13}{3} \binom{13}{3}}
=\frac{3 \binom{13}{4}  \binom{13}{2}}{ \binom{13}{3} \binom{13}{3}}$$

* Expand binomial coefficients in terms of factorials, and cancel common factors.
$$\frac{3 \cdot 13! 13! 10! 3! 10! 3!}{9! 4! 11! 2! 13! 13!} = \frac{3 \cdot 10! 3! 10! 3!}{9! 4! 11! 2!}$$

* Cancel nearby factorials from numerator and denominator
$$3 \cdot \frac{10!}{9!} \frac{10!}{11!} \frac{3!}{4!} \frac{3!}{2!} = \frac{3 \cdot 10 \cdot 3}{11 \cdot 4} = \frac{45}{22}$$


# Overcounting and correction

* It is OK to overcount, as long as you know by what factor you are overcounting and divide by that factor.  

* Example 1: A regular dodecahedron, as its name suggests, has 12 faces, each of which is a regular pentagon. Three faces meet at each vertex. How many vertices does the dodecahedron have?

* Wrong: 12 pentagons $\times$  5 vertices per pentagon = 60.  

* Correct: Since each vertex is part of three pentagons, we overcounted by a factor of 3. The correct answer is 60/3 = 20.



* Example 2: How many different strings can you make by rearranging the letters in the word "SUCCESS"?

* Wrong: The number of permutations of 7 symbols, 7!, would count each string 12 times, because there are 6 ways to reorder SSS and two ways to reorder CC.

* The correct answer is $$\frac{7!}{3!2!} = 420.$$

# Counting arrangements of indistinguishable objects

* If you have two different types of object in a sequence, $n$ of type 1 and $k$ of type 2,  but the objects of each type are indistinguishable, the number of arrangements is equal to the number of ways of selecting the set of positions of the type 2 objects:
$$
\binom{n+k}{k}=\frac{(n+k)!}{n! k!}.
$$


* Example: a football team ends the regular season with an 11-5 record. How many different sequences of wins and losses can lead to this outcome?

* Answer: There are $\binom{16}{5}$ ways to place the five losses among the 16 games. Equivalently, there are $\binom{16}{11}$ ways to place the eleven wins among the 16 games.

# More examples of the divider strategy 
* If there is a bijection between two sets $A$ and $B$, they have the same cardinality. Sometimes it is easier to count one than the other.

* Example: You can divide a set of $n$ objects into $k$ subsets by inserting $k-1$ dividers. For example, to find all six-element strings that begin with 0 or more A's, continue with 0 or more B's, and end with 0 or more C's, just put one divider between the last A and the first B, a second divider between the last B and the first C. The bijection maps a pattern of letters and dividers like $xxx|x|xx$ into a string like AAABCC. Then there are $\binom{6+2}{2}$ different ways to place the two dividers.


* Another example:
How many different terms of the form $x^ix^jx^kx^l$ are equal to $x^{11}?$  
Clever trick: Each term is like $xxx|xxxxx|x|xx$.

* So there are $\binom{11+3}{3}$ ways to place the dividers.  
Notice that ||xxxxxxxxxxx| coresponds to $x^0x^0x^{11}x^0$.

#  Finitely additive set functions

* We consider functions $f: \mathcal{F} \rightarrow \mathbb{R}$, where $\mathcal{F}$ is a collection of some (perhaps all) of the subsets of a "universal set'' $S.$ The set $S$ can be finite, countably infinite, or uncountably infinite. 

* Example: $\mathcal{F}$ is a collection of regions (possibly overlapping) in the United States, and $f$ specifies the area or population of a region.

* For $A, B \in \mathcal{F}$, $A - B$ denotes the set of elements of $S$ that are in $A$ but not in $B$, and $A^c$, the complement of $A$, denotes $S-A$.

* $A$ and $B$ are **disjoint** if they have no elements in common: $A \cap B = \emptyset.$ 

* Function $f$ is **finitely additive** if, for disjoint sets,   $f(A \cup B) = f(A) + f(B).$ 

# Inclusion-exclusion

* The inclusion-exclusion principle extends this result to sets that are not disjoint. It asserts that for a finitely additive set function,  
$f(A \cup B) = f(A) + f(B)-f(A \cap B).$

* The secret of the proof is to write sets as disjoint unions of other sets so that you can apply the additivity property.

  + $f(A \cup B) = f(A) + f(B-A)$ because $(A \cup B) = A \cup (B-A)$ (disjoint)
  
  + $f(B) = f(A\cap B) + f(B-A)$ because $B = (A\cap B) \cup (B-A)$ (disjoint)
  
* Subtract to find $f(A \cup B) - f(B) = f(A) - f(A\cap B)$
and rearrange to get $f(A \cup B) = f(A) + f(B)-f(A \cap B).$

* Nothing in this proof required $f$ to be nonnegative.

# Simple application of inclusion-exclusion

* Over the last 24 hours, my high-tech stocks are up by \$40. My high-tech domestic stocks are up by \$100, while my stocks that are either high-tech or domestic are down by \$40.
What was the change in value of my domestic stocks?

* Solution: Let $T$ be the subset of high-tech stocks, let $D$ be the subset of domestic stocks. Then  
$f(T) = +40, f(T \cap D) = +100, f(T \cup D) = -40$

* Inclusion-exclusion says that $f(T \cup D) = f(T) + f(D)-f(T \cap D).$  
So $-40 = 40 + f(D) -100$ and $f(D) = 100 -40 -40 = 20.$  
Negative numbers posed no problem. 
 
# Finitely additive measure

* If a finitely additive set function is non-negative, it is called a **finitely additive measure**. From this property we can deduce inequalities like
$f(A \cup B) \leq f(A) + f(B).$
 
* Proof: $f(A \cup B) = f(A) + f(B)-f(A \cap B)$   but now $f(A \cap B) \geq 0$ and so   
$f(A \cup B) \leq  f(A) + f(B)$.

# Probability measure

* If $f$ has the property that $f(S) = 1$, it is called a **finite probability measure** and is denoted $\mathbf{P}(A)$. In this case, a subset $A \subset S$ is called an "event". 

* Now we can make statements like
$\mathbf{P}(A^c) = 1 - \mathbf{P}(A).$  
 
* Proof: $A \cup A^c = S$ (disjoint) so $\mathbf{P}(A) + \mathbf{P}(A^c) = 1.$  





# Random variables and expectation

* A **random variable** is not a variable! It is a function $X: S \rightarrow \mathbb{R}.$ If $S$ is finite, the image of $X$ is a finite subset of $\mathbb{R}$. 

* A **probability space** $(S,\mathcal{F},\mathbf{P})$ requires a sample space, a sigma-field, and a probability measure.

* The **expectation** of a **simple** random variable $X$, one which can assume only a finite number $k$ of different values $c_i$, is defined as a sum over those values:
$$E(X) = \sum_{i=1}^k c_i \mathbf{P}(X = c_i).$$

* In order to define the expectation of $X$, we impose the requirement that $X$ is "measurable": the event "$X$ has the value $c_i$" is in the sigma-field $\mathcal{F}$.  

* Equivalently, we could have imposed the more general requirement that the event $X \leq x$ is in the sigma-field for any real number $x$. However, that would lead to an unnecessarily large sigma-field: who cares whether the total roll on two dice is $\leq 2 \pi?$


* The definition of expectation can work even if the sample space is not finite! For example, $S$ could be $[0, 100]$, with $X(s) = 1$ if $s \leq 20$,   $X(s) = 2$ if $20 < s \leq 60$, $X(s) = 3$ if $s > 60$.  
But now we need a bigger sigma-field, with events like $20 < s \leq 60$.

# Indicator functions

* The **indicator function** $\mathbf{1}_A$, which has the value 1 if $\omega \in A$, 0 otherwise, is a random variable on the probability space.  

* Then for any event $A$,  $E(\mathbf{1}_A) =\mathbf{P}(A)$.  
Proof: $E(\mathbf{1}_A) = 1 \cdot \mathbf{P}(\mathbf{1}_A = 1) + 0 \cdot \mathbf{P}(\mathbf{1}_A = 0) = 1 \cdot \mathbf{P}(A) = \mathbf{P}(A)$

# Random variables in terms of indicator functions

* Any simple random variable $X$ can be expressed uniquely in the form
$X = \sum_{i=1}c_i.\mathbf{1}_{A_i} .$
where the values $c_i$ are the values that can be assumed by $X$ and the events $A_i$ are the inverse images $X^{-1}({c_i})$, for which $X(s) = c_i$ if $s \in A_i.$
 
* Then $E(X)= \sum_{i=1}c_i\mathbf{P}(A_i) = \sum_{i=1}^n c_i \mathbf{P}(X = c_i).$
 
* If $S$ is finite (size $n$), we can use the sigma-field $\mathcal{F} = 2^S$, although it may be larger than the sigma-field generated by the events $A_i$.

* Then any single-element set $\{s\}$ is in the sigma-field and we can define an indicator function $\mathbf{1}_{\{s\}}$ for it.   
 The indicator function for the event $A_i$ can now become a sum over the sample space:
$$\mathbf{1}_{A_i} = \sum_{s: X(s) = c_i}\mathbf{1}_{\{s\}},\text{   and}$$
$$E(X) = \sum_{i=1}^kc_i\mathbf{1}_{A_i} = \sum_{i=1}^kc_i  \sum_{s: X(s) = c_i}\mathbf{1}_{\{s\}}= .\sum_{s \in S}X(s)\mathbf{1}_{\{s\}} $$
$$\text{ or   }E(X) = \sum_{i=1}^kc_i\mathbf{P}(A_i) = \sum_{s \in S}X(s)\mathbf{P}(\{s\}).$$
This is a special case of the so-called "law of the unconscious statistician.'' 



# Working with the unconscious statistician:

You roll a single fair die. If the roll is 1, your payoff $X$ is 6. For a larger odd roll, your payoff $X$ is 12.  For an even roll, your payoff $X$ is 18. There are two ways to express $X$ as a sum of indicator functions:

* Over the image of $X$.  
$X = 6 \cdot \mathbf{1}_{X=6} + 12 \cdot \mathbf{1}_{X=12} + 18 \cdot \mathbf{1}_{X=18}$
 
* Over the domain of $X$ (the sample space)  
$X = 6 \cdot \mathbf{1}_{1} + 12 \cdot \mathbf{1}_{\{3,5\}} + 18 \cdot \mathbf{1}_{\{2,4,6\}}$
 
* Here is a calculation of $E(X)$ that makes direct use of the definition of expectation. There is one term for each possible value of $X$.  
$E(X) = 6 \cdot \mathbf{P}(X = 6) + 12 \cdot \mathbf{P}(X = 12)+ 18 \cdot \mathbf{P}(X = 18)$  
$E(X) = 6 \cdot \frac{1}{6} +  12 \cdot \frac{1}{3} +  18 \cdot \frac{1}{2} = 1+4+9 = 14$

* Here is a calculation of $E(X)$ that uses the law of the unconscious statistician. There is one term for each point in the domain.  
$E(X) = 6 \cdot \mathbf{P}(1) + 18 \cdot \mathbf{P}(2)+ 18 \cdot \mathbf{P}(3) +18 \cdot \mathbf{P}(4) + 12 \cdot \mathbf{P}(5)+ 18 \cdot \mathbf{P}(6)$  

* $E(X) = \frac{1}{6}(6 + 18 +12 +18+12+18) = \frac{84}{6} = 14.$



# The sum of two simple random variables
* Suppose that random variable $X$ takes values in the finite set $x_i$ and random variable $Y$ takes values in the set $y_i$. In order to define a probability function on the sample space $S$, we must know the probability of each event of the form $(X = x_i, Y = y_j).$

* If $X$ and $Y$ are simple random variables, each of which takes on values in a finite subset of $\mathbb{R},$ and $Z = X+Y$, then $E(Z) = E(X) + E(Y)$.

* Proof: $E(Z) = E(X+Y) = \sum_i \sum_j (x_i +y_j) \mathbf{P}(X = x_i, Y = y_j)$

* Split the sum: $$E(Z) = \sum_i \sum_j x_i \mathbf{P}(X = x_i, Y = y_j) + \sum_i \sum_j y_j \mathbf{P}(X = x_i, Y = y_j)$$

* In each term, one of the sums is just a sum over all values of one variable:  
$$E(Z) = \sum_i (x_i) \mathbf{P}(X = x_i) +  \sum_j (y_j) \mathbf{P} (Y = y_j) = E(X) + E(Y)$$

# The product of two independent simple random variables

* Random variables $X$ and $Y$ are called  **independent** if   
$\mathbf{P}(X = x, Y = y) = \mathbf{P}(X=x)\mathbf{P}(Y=y)$.   
If $X$ and $Y$ are **independent** simple random variables and $W = XY$, then $E(W)= E(X) E(Y)$. 

* Proof: $$E(W)= E(XY) = \sum_i \sum_j x_i y_j \mathbf{P}(X = x_i, Y = y_j).$$

* Exploit independence: $E(W)= \sum_i \sum_j x_i y_j \mathbf{P}(X = x_i), \mathbf{P}(Y = y_j)$.

* Factor: $E(X) = \sum_i x_i \mathbf{P}(X = x_i) \sum_j y_j \mathbf{P}(Y = y_j) = E(X)E(Y)$


# Uncorrelated does not necessarily mean independent

* The converse of the result just proved is not necessarily true: it is possible to invent "uncorrelated'' random variables, for which $E(XY) = E(X)E(Y)$, that are not independent.

* Example: You take a two-question true-false test, guessing on each question so that your probability of bing correct is 1/2. "Right" scores +1, "wrong" scores -1. Random variable $X$ is your total score , -2, 0, or +2.
Random variable $Y$ is your improvement from the first question to the second, -2, 0, or +2.
$$
\begin{array}{|c|c|c|c|c|c|}
\hline
Probability & Q1 Score  & Q2 Score & X & Y & W = XY\\
\hline 
1/4 & 1 & 1  & +2 & 0 & 0 \\
\hline 
1/4 &1& -1  & 0 & -2  & 0\\
\hline 
1/4 &-1 & 1  &  0  & +2 & 0\\
\hline 
1/4 &-1& -1 & -2  & 0 & 0\\
\hline 
\end{array}
$$

* Now $E(X) = 0, E(Y) =0, E(XY) = 0$.  
So $E(XY) = E(X)E(Y)$ and $X$ and $Y$ are uncorrelated.

* But $X$ and $Y$ are not independent. One counterexample suffices to show this:   
$\mathbf{P}(X=2, Y=2) = 0$ but  
$\mathbf{P}(X=2) \mathbf{P}(Y=2) = \frac{1}{4} \cdot \frac{1}{4} \neq 0$

# Definition and calculation of variance

* The variance Var$(X)$ of a random variable $X$ is defined as $E((X - E(X))^2) .$

* Given that $E(a_1X_1+a_2X_2) = a_1E(X_1) + a_2E(X_2)$ in all cases and that   $E(X_1X_2) = E(X_1) E(X_2)$ for independent random variables, we can prove three key results about variance.

# One-pass calculation of variance

* First note that $E(X)$ is a constant and so $E(E(X)) = E(X).$ 

* Expand: $Var(X) = E((X - E(X))^2) = E(X^2 - 2E(X) \cdot X + E(X)^2)$

* Evaluate:  $Var(X) = E(X^2) - 2E(X) \cdot E(X) + E(X)^2)$

* Simplify:  $Var(X) = E(X^2) - E(X)^2$

*  In a single pass through the sample space we can compute both $E(X)$ and $E(X^2).$

# The effect of multiplying by a constant $a$ and adding a constant $b$

* Use the previous formula: Var$(aX+b) = E((aX+b)^2) - (E(ax+b))^2$

* Expand: Var$(aX+b) = E(a^2X^2) + 2E(ax)b + b^2 - E(aX)^2 - 2 E(aX)b - b^2$

* Cancel and simplify: Var$(aX+b) = a^2E(X^2) - a^2(E(X)^2) = a^2 Var(X)$

# Variance of the sum of two independent random variables

* We will exploit the fact that in this case $E(XY) = E(X)E(Y)$. The variables $X$ and $Y$ only need to be uncorrelated, not necessarily independent.

  + Use the one-pass formula: $Var(X+Y) = E((X+Y)^2) - E(X+Y)^2$
  
  + Expand: $Var(X+Y) = E(X^2) +2E(xY) +E(Y^2) -E(X^2) - 2E(X)E(Y) - E(Y)^2$
  
  + Combine terms: $Var(X+Y) = Var(X) + Var(Y) +2(E(XY) - E(X)E(Y))$

* If $X$ and $Y$ are independent, they are uncorrelated and the last term is zero, in which case
$Var(X+Y) = Var(X) + Var(Y)$

* This result is easily extended by induction to the case of $n$ random variables that are "pairwise uncorrelated" in the sense that for $i \neq j$, $E(X_iX_j) = E(X_i)E(X_j)$. 

* In the special case where $X$ is the sum of $n$ independent random variables, each with the same variance $\sigma^2$,
$$Var(X) = n \sigma^2.$$

# Variance of the mean of $n$ independent random variables

* This is a theorem of probability.  
Suppose that $X_1, X_2, \cdots X_n$ are independent random variables, all with the same expectation $\mu$ and variance $\sigma^2.$   
Their mean is $\overline{X} = \frac{1}{n}(X_1 + X_2 + \cdots X_n).$

* Then $$E(\overline{X}) = \frac{1}{n}(E(X_1)+ \cdots + E(X_n)) = \frac{1}{n}(n\mu) = \mu$$

* $$Var(\overline{X}) = (\frac{1}{n})^2(Var(X_1)+ \cdots + Var(X_n)) = \frac{1}{n^2}(n \sigma^2) = \frac{1}{n} \sigma^2$$

# Variance of the sample mean of $n$ independent random variables.

* This is a theorem of statistics.  
Let $X_1, X_2, \cdots X_n$ be independent random variables, each from a distribution with  
Var$(X_i) = \sigma^2 < \infty.$

* The **sample mean** is $\overline{X} = \frac{1}{n}(X_1 + X_2 + \cdots X_n).$

* We do not know the expectation $\mu=E(x_i)$, although we know from the previous result that the **expectation** of $\overline{X}$ is equal to $\mu$.

* We also do not know $\sigma^2$, the variance of $X_i$. We can try to estimate it by using the usual formula but, not knowing $\mu$, we can do no better than to use $\overline{X}$ in its place. The challenge:
$\text{Prove that }E(\sum_{i=1}^n(X_i - \overline{X})^2)=(n-1)\sigma^2.$

# Proof for the variance of the sample mean
* Preliminary algebra: $$\sum_{i=1}^n(X_i - \overline{X})^2 = \sum_{i=1}^nX_i^2 - 2 \overline{X}\sum_{i=1}^nX_i+n \overline{X}^2 $$
$$\sum_{i=1}^n(X_i - \overline{X})^2 = \sum_{i=1}^nX_i^2 - 2n \overline{X}^2 +n \overline{X}^2 =  \sum_{i=1}^nX_i^2-n \overline{X}^2.$$

* Now take expectations: $$E(\sum_{i=1}^n(X_i - \overline{X})^2)  = E(\sum_{i=1}^nX_i^2) -n E(\overline{X})^2.$$
 
* Use the formula $Var (X_i) = \sigma^2 = E(X_i^2) - \mu^2$ backwards to get 
$$E(\sum_{i=1}^n(X_i - \overline{X})^2) = n(\sigma^2 + \mu^2) - n(Var(\overline{X}) + \mu^2)$$
$$E(\sum_{i=1}^n(X_i - \overline{X})^2) = n(\sigma^2 + \mu^2) - n(\frac{\sigma^2}{n} + \mu^2)$$
$$E(\sum_{i=1}^n(X_i - \overline{X})^2) = n\sigma^2 - \sigma^2 = (n - 1) \sigma^2$$

# Estimating the variance of a random variable from sample variance
* We have found that $$E(\sum_{i=1}^n(X_i - \overline{X})^2) = (n - 1) \sigma^2$$

* If we have measured the sample variance, the best estimate of the unknown variance $\sigma^2$ for an individual measurement in a sample is $$S^2 = \frac{1}{n-1} \sum_{i=1}^n(X_i - \overline{X})^2.$$ This will not equal $\sigma^2$, but it has the nice property of being an "unbiased" estimate whose **expectation** is equal to $\sigma^2.$ 

* This \emph{sample variance} $S^2$ is what ```var()``` computes in R. It is usually appropriate when you are dealing with real data, since you probably have a sample from a larger population and want to estimate the variance for the population. If, however, you model a **population** by constructing a dataframe with one row for each equally likely element, do not use ```var()```.   Here is a simple example in R:
```{r, echo = TRUE}
v <- 1:6; v #outcomes for roll of a single die
mu <- mean(v); mu #the expectation of the roll
sigsq <- mean(v^2) - mu^2; sigsq #the variance of the population
var(v)   #this is wrong - it assumes we had a sample from a larger population
#Let's confirm this by sampling from the population.
#First do a single sample
x <- sample(v, 5, replace = TRUE)  # sample of five drawn from this population
x.bar <- mean(x); x.bar   #the sample mean - a random variable
x.Var <- mean(x^2)-x.bar^2; x.Var  #the variance of the numbers in the sample - we divided by 5
Ssq <- (1/4)*(sum(x^2) - 5*x.bar^2); Ssq #the sample variance
var(x)  #the built-in function gives the sample variance
#Now repeat with 10000 samples
N <- 10000; vecMu <- vecVar <- vecSsq <- numeric(N); #create empty vectors
for (i in 1:N){
  x <- sample(v, 5, replace = TRUE)  # sample of five drawn from this population
  vecMu[i] <- mean(x)   #the sample mean - a random variable
  vecVar[i] <- mean(x^2)-mean(x)^2  #the variance of the numbers in the sample - we divided by 5
  vecSsq[i] <- (1/4)*(sum(x^2) - 5*mean(x)^2) #the sample variance
}
hist(vecMu, breaks = 25)
mean(vecMu); mu   #agrees with the population mean
hist(vecVar, breaks = 25)
mean(vecVar); sigsq   #does not agree with the population variance
mean(vecSsq); sigsq   #agrees with the population variance
```

# Using R to do finite probability

* A **data frame** in R is a list of vectors, all of the same length, placed side-by-side in a two-dimensional array.

* The sample space $S$ consists of all the rows in a data frame.

* An outcome $\omega$ of an "experiment'' is an individual row.

* An event $A$ is just a subset of the rows. 

* The sigma field of events to which probabilities can be assigned is the power set $2^S$: the collection of all possible subsets. 

* For a probability function, assume that each of the $n$ rows has probability $1/n$. Then the probability of event $A$ can be determined by counting the number of rows in event (subset) $A$.

* If you want unequal probabilities whose ratio is a rational number like $3/2$, include three rows for the first outcome, two for the second.

# Numeric, logical and factor columns
* The difference between a matrix and a dataframe in R is that the columns of a matrix must be **numeric** vectors, while the columns of a dataframe may be numeric, logical, or factor columns.

* A numeric column in a data frame is a random variable, since it assigns a real number to each row. When the "vectorized'' operators in R act on columns of a data frame, they carry out the specified operation on each row and  can be viewed as operations on random variables.

* A constant, since it gets replicated $n$ times when combined with a vector of length $n$, behaves like a constant random variable when it appears in an $R$ expression like `X+2`.

* A logical column in a data frame specifies an event $A$, the set of rows for which the value is `TRUE`.  The complement of event $A$ is the the set of rows for which the value is `FALSE`. 

* Since `TRUE` has the value 1, a logical column can be treated like a random variable; namely, the "indicator function" that equals 1 if event $A$ occurs, 0 if it does not.

* A "factor" column in a data frame specifies a **partition** of the sample space into disjoint events (subsets) $B_1, B_2, \cdots B_k.$  

# Calculating means and variances from a data frame
* The R function `mean()` computes the expectation of a random variable.

* The R function `var()` does not compute the variance of a random variable. It assumes that it has been given a vector of samples, and it estimates the variance of the population from which those samples were drawn.

* If you use a dataframe to calculate the variance of a random variable X, use `mean(X^2) - mean(X)^2`.

* Many elementary probability problems can be solved by counting how many rows in a data frame correspond to a specified event.

# Example with unequal probabilities
* On a fair roulette wheel with no 0, you can bet that the number that arises is divisible by 3. If it is, you win 2 chips, if not, you lose 1 chip. Suppose you place five such bets. What are the expectation and variance of your net winninngs, and what is the probability of winning two or more times?
```{r, echo = TRUE}
play1 = c(2,-1,-1)  #the equally likely outcomes for 1 play
df <- expand.grid(play1, play1, play1, play1, play1); head(df)
#Create a new random variable for the net winnings
df$net <- df$Var1 + df$Var2 +df$Var3 + df$Var4 +df$Var5; head(df)
mean(df$net)  #no surprise -- the wheel was fair
mean(df$net^2) - mean(df$net)^2   #the variance
#There is a formula npq for the variance in this case of a binomial distribution
#but we must multiply by 3^2 = 9
9 * 5 * (1/3) * (2/3)   #variance as calculated from the binomial distribution
var(df$net)   #wrong, but with so many rows it makes little difference
mean(df$net > 0) #probability of 2 or more wins
```

# The built-in binomial distribution

* In the previous example, we can calculate the probability of $k$ wins in 5 spins
```{r}
n <- 5   #number of spins of the wheel
p <- 1/3   #probability of winning on a single spin
pWin <- function(k) choose(n,k)*p^k*(1-p)^(n-k)
pWin(2)
#R has a built-in function for this calculation
dbinom(2,n,p)
#It can also tell you the probability of more than 1 win
pbinom(1,n,p, lower.tail = FALSE); 0.5390947  #compare with previous number
# Making a barplot requires clever use of outer()
tbl <- outer(0:n, p, dbinom, size = n); tbl
#This is almost right but we need to transpose it!
barplot(t(tbl), names.arg = 0:n)

```


# A standard application of inclusion-exclusion, done by brute force in R

* What is the probability, when two different cards are selected at random from a deck, that at least one is a spade?
```{r, echo = TRUE}
df <- expand.grid(1:52, 1:52); head(df)
#Eliminate rows with duplicate cards
df <- subset(df, !(Var1==Var2)); nrow(df); 52*51
#If the card number is divisible by 4, the card is a spade
mean((df$Var1%%4 == 0)| (df$Var2%%4 == 0))
1 - (39*38)/(52*51)   #the conventional way to get the answer
```
# How to conduct a permutation test, as in script 2D

* This single technique can replace much of elementary classical statistics.

  + You have $m$ values for one subgroup (ounces of beer consumed by males)
  + You have $n$ values for the other subgroup (ounces of beer consumed by females)
  + You have computed the mean values for the two subgroups, and they are different.
  + You want to assess whether this difference is significant.
  + The strategy is to determine how frequently such a large difference is likely to arise if, instead of the $m$ males, you use a random subset of $m$ of the $m+n$ beer drinkers.


* Here is the algorithm, taken verbatim from Chihara and Hesterberg,
\underline{Mathematical Statistics with Resampling and R}

* Pool the $m+n$ values.

* **repeat**  
  + Draw a resample of size $m$ without replacement.
  + Use the remaining $n$ observations for the other sample.
  + Calculate the difference of means or another statistic that compares samples,
**until** we have enough samples.

* Calculate the P-value as the fraction of times the random statistics exceed the original statistic. Multiply by 2 for a 2-sided test.

* Optionally, plot a histogram of the random statistic values.

# Permutation test for beerwings
```{r, echo = TRUE}
#Load the beerwings dataset from a file in the current directory (make sure it's the default)
bw <- read.csv("beerwings.csv"); head(bw)
#Divide beer consumed by men by the number of men
beerMale <- sum(bw$Beer*(bw$Gender=="M"))/sum(bw$Gender=="M"); beerMale
#Divide beer consumed by women by the number of women
beerFemale <- sum(bw$Beer*(bw$Gender=="F"))/sum(bw$Gender=="F"); beerFemale
beerDiff <- beerMale-beerFemale; beerDiff  #is this difference significant?

#Repeat with a random sample, same size as the subset of men
samp <- sample(nrow(bw), sum(bw$Gender == "M")); samp
beerSamp <- mean(bw$Beer[samp]); beerSamp; #mean for the random sample
beerOther <- mean(bw$Beer[-samp]); beerOther; #mean for the complement of the random sample
diff <- beerSamp - beerOther

#Once it works, do it 10000 times
N <- 10000; diff <- numeric(N)
for (i in 1:N) {
  samp <- sample(nrow(bw), sum(bw$Gender == "M"))
  beerSamp <- mean(bw$Beer[samp]) #mean for the random sample
  beerOther <- mean(bw$Beer[-samp]); #mean for the complement of the random sample
  diff[i] <- beerSamp - beerOther
}
hist(diff, breaks = 30)
abline(v= beerDiff, col = "red") #observed difference is way out on the tail
mean(diff >= beerDiff)  #Usually the P-value is about 3%
#The textbook does hotwings, for which the P-value is smaller
```
# Munchkin problems - the hypergeometric distribution

* This is based on a true story.  
To placate her preschoolers while waiting for a flight from Logan Airport, Lisa buys six Dunkin' Muchkins, four plain and two chocolate. She chooses three at random and puts them in a bag for Thomas. The remaining three go into a bag for Catherine.  
Find the probability for Thomas to have 0, 1, or 2 chocolate Munchkins.

* Solution: there are $\binom{6}{3} = \frac{6!}{3!3!}= \frac{6\cdot 5 \cdot 4}{3\cdot 2 \cdot 1}=20$ ways to select the Munchkins for Thomas.  
The number of ways to select $k$ chocolate Munchkins is $\binom{2}{k}$$\binom{4}{3-k}$.  

* If $k=1$ the probability is $$\frac{\binom{2}{1}\binom{4}{2}}{20}=\frac{2\cdot 6}{20}= \frac{3}{5}.$$

* If $k=2$ (or 0) the probability is $$\frac{\binom{2}{2}\binom{4}{1}}{20}=\frac{1\cdot 4}{20}= \frac{1}{5}.$$

* The number $k$ of Munchkins in Thomas's bag is said to have a **hypergeometric distribution.**

# Permutation Tests and Contingency Tables

The Munchkin problem can be solved by making two columns for the positions of the chocolate Munchkins.

```{r}
#install.packages("combinat")
library(combinat)
positions <- combn(1:6,2); positions #a matrix with two rows
choc1 <- positions[1,]
choc2 <- positions[2,]
df <- data.frame(choc1, choc2); head(df)
#Thomas gets the Munchkins in positions 1, 2, and 3
sum(((choc1 < 4)+(choc2<4)) == 0)/nrow(df)
sum(((choc1 < 4)+(choc2<4)) == 1)/nrow(df)
sum(((choc1 < 4)+(choc2<4)) == 2)/nrow(df)
```

# Permutation test for Munchkins

* We can use our standard permutation test approach.
```{r}
 m <- 2  #number of type 1 objects (chocolate Munchkins)
 n <- 4  #number of type 2 objects (plain Munchkins)
 k <- 3  #number of Munchkins in Thomas's bag
 Thomas <- c(rep(TRUE,k),rep(FALSE,m+n-k))   #ownership
 Munchkins <- c(rep(TRUE,2),rep(FALSE,4))   #flavor
 #Instead of making a row for each possible distribution and counting,
 #we do a simulation.
 N <- 10000; TC <- numeric(N)   #number of chocolate in Thomas's bag
 for (i in 1:N){
   scramble <- sample(Munchkins, m+n, replace = FALSE)  #permute the Munchkins
   TC[i] <- sum(Thomas&scramble)
 }
 table(TC)/N    #will be close to the correct probability
 #We can make a contingency table from our fabricated data
 table(Thomas,Munchkins)
 #It's too small for chi-square, but R can replicate our exact analysis
 fisher.test(Thomas,Munchkins, alternative = "g")   #Probability of 2 or more chocolate
 # Put just one Munchkin in Thomas's bag
 Munchkins <- c(TRUE, FALSE, FALSE, TRUE, FALSE, FALSE)
 fisher.test(Thomas,Munchkins, alternative = "g")   #Probability of 1 or more chocolate
 dhyper(0:2, m,n,k)   #mass function for the hypergeometric distribution
 1 - phyper(0:2,m,n,k)  #probability of more than 0, 1, 2
 
 
 # This approach is better than chi-square for 2x2 tables.
 
 GSS <- read.csv("GSSLogical.csv"); head(GSS);nrow(GSS)
 attach(GSS)
 table(Protestant, Republican)
 m <- 239  #number of type 1 objects (Republican)
 n <- 316  #number of type 2 objects (Democrats)
 k <- 256  #number of Protestants
 N <- 10000; RP <- numeric(N)   #number of Republican Protestants
 for (i in 1:N){
   scramble <- sample(Republican, m+n, replace = FALSE)  #permute the politics
   RP[i] <- sum(Protestant&scramble)
 }
 hist(RP, breaks = 30)
 mean(RP > 138)  #probability of 139 or more Protestant Republicans
 fisher.test(Republican,Protestant, alternative = "g")
 chisq.test(Republican,Protestant)   #this does a two-sided test
```



















