---
title: "Logic, using Boolean Algebra and Finite Fields"
author: "Arnob Roy"
date: "Mathematics for Computation and Data Science"
output: pdf_document
---

#

# R Homework Problems
## Problem 1. Truth Tables in R
### 1.a. $(p\lor q)\implies (r\land s)$

```{r}
p <- c(rep(TRUE,8), rep(FALSE,8));p
q <- rep(c(T,T,T,T,F,F,F,F),2);q
r <- rep(c(T,T,F,F),4);r
s <- rep(c(T,F),8);s
```

Consider the logical expression $(p\land q)\to (r\lor s)$, as polynomial expression in finite field $\mathbb{Z}_2, 1+(pq)+(pq)(r+s+rs).$ This expression simplifies:

\[1+(pq)+(pq)(r+s+rs) = 1+pq+pqr+pqs+pqrs\]

```{r}
"%+%" <- function(x,y) (x+y)%%2
IMPLIES <- 1 %+% (p*q) %+% (p*q*r) %+% (p*q*s) %+% (p*q*r*s)
IMPLIES2 <- 1 %+% (p*q) %+% ((p*q)*(r %+% s %+% (r*s)))
data.frame(p,q,r,s,IMPLIES, IMPLIES2)
```

### 1.b. Write the expression in conjunctive normal form using the truth table

(ANDs of OR-terms, with each variable included in every OR-term)

Using line 4, we have $(\neg p \lor \neg q \lor r \lor s).$

### 1.c. Generate the truth table

```{r}
data.frame(p,q,r,s, (!p|!q|r|s))
```


\newpage

## Problem 2. Data Wrangling, Descriptive Statistics, and Chi-Square Tests

### 2.a. Female Democrat Gun Owners

```{r}
data2a <- read.csv("GSSLogical.csv")
Problem2a <- subset(data2a, Male==FALSE&Republican==FALSE&GunOwner==TRUE, c(Male,Republican,GunOwner)); summary(Problem2a)
```

I observed that there were 37 female non-Republican gun owners.

### 2.b. Games with over 10 home-runs!
```{r}
data2b <- read.csv("RedSox2013.csv")

Problem2b <- data.frame(data2b$Game, data2b$R, data2b$RA, totalscore=(data2b$R+data2b$RA), duration=data2b$Duration)

Problem2bsubset <- subset(Problem2b,totalscore>=10 )

summary(Problem2bsubset)
```
I observed that the shortest duration in which the two teams scored over 10 home-runs together was 153 (I'd guess the units are in minutes).

### 2.c. Visualizing the number of home-runs using Boxplots
```{r}
data2c <- read.csv("RedSox2013.csv")

Problem2cDay <- subset(data2c, DayNight == "D" )
Problem2cNight <- subset(data2c, DayNight == "N" )

boxplot(Problem2cDay$R, xlab = "Day Time Game", ylab = "Number of Home-runs", col = "Light blue")
boxplot(Problem2cNight$R, xlab = "Night Time Game", ylab = "Number of Home-runs", col = "Dark Blue")
```

### 2.d. Creating Tables using Outer() & the Chi-Square Test

```{r}
data2d <- read.csv("RedSox2013.csv"); data2d

tbl1 <- table(data2d$WonLost, data2d$Away); tbl1
Expected <- outer(rowSums(tbl1), colSums(tbl1))/sum(tbl1); Expected
chisq.test(data2d$WonLost, data2d$Away)
```

```{r}
tbl2 <- table(data2d$WonLost, data2d$DayNight); tbl2
Expected <- outer(rowSums(tbl2), colSums(tbl2))/sum(tbl2); Expected
chisq.test(data2d$WonLost, data2d$DayNight)
```

\newpage

## Problem 3. Examining Air Quality, Boxplots, and Chi-Square Tests


```{r}
data("airquality")
summary(airquality)

May <- subset(airquality, Month==5, na.rm=TRUE)
boxplot(May$Ozone, xlab = "May", ylab = "Ozone")
boxplot(May$Solar.R, xlab = "May", ylab = "Solar Radiation")
boxplot(May$Wind, xlab = "May", ylab = "Wind")
boxplot(May$Temp, xlab = "May", ylab = "Temperature")

Jun <- subset(airquality, Month==6, na.rm=TRUE)
boxplot(Jun$Ozone, xlab = "June", ylab = "Ozone")
boxplot(Jun$Solar.R, xlab = "June", ylab = "Solar Radiation")
boxplot(Jun$Wind, xlab = "June", ylab = "Wind")
boxplot(Jun$Temp, xlab = "June", ylab = "Temperature")

Jul <- subset(airquality, Month==7, na.rm=TRUE)
boxplot(Jul$Ozone, xlab = "July", ylab = "Ozone")
boxplot(Jul$Solar.R, xlab = "July", ylab = "Solar Radiation")
boxplot(Jul$Wind, xlab = "July", ylab = "Wind")
boxplot(Jul$Temp, xlab = "July", ylab = "Temperature")

Aug <- subset(airquality, Month==8, na.rm=TRUE)
boxplot(Aug$Ozone, xlab = "August", ylab = "Ozone")
boxplot(Aug$Solar.R, xlab = "August", ylab = "Solar Radiation")
boxplot(Aug$Wind, xlab = "August", ylab = "Wind")
boxplot(Aug$Temp, xlab = "August", ylab = "Temperature")

Sep <- subset(airquality, Month==9, na.rm=TRUE)
boxplot(Sep$Ozone, xlab = "September", ylab = "Ozone")
boxplot(Sep$Solar.R, xlab = "September", ylab = "Solar Radiation")
boxplot(Sep$Wind, xlab = "September", ylab = "Wind")
boxplot(Sep$Temp, xlab = "September", ylab = "Temperature")

```

### b. Scatterplots

Simply examining the plots below, I saw that Temperature may not have a strong association with Solar Radiation; however, I would recommend considering confounding variables before making a final judgment. Then, I observed that temperature may have a positive association with Ozone, as the amount of Ozone

```{r}
plot(airquality$Temp, airquality$Solar.R, xlab = "Temperature", ylab = "Solar Radiation")
plot(airquality$Temp, airquality$Ozone, xlab = "Temperature", ylab = "Ozone")
```

### c. Defining and using logical vectors in a dataset

Below listed is the subset of the dataset which includes days for which all four quantities were higher than the average for that quantity.

```{r}
airquality$smoggy <- airquality$Ozone > mean(airquality$Ozone, na.rm=TRUE)
airquality$windy <- airquality$Wind > mean(airquality$Wind, na.rm=TRUE)
airquality$sunny <- airquality$Solar.R > mean(airquality$Solar.R, na.rm=TRUE)
airquality$hot <- airquality$Temp > mean(airquality$Temp, na.rm=TRUE)

airquality$allfour <- (airquality$smoggy & airquality$windy & airquality$sunny & airquality$hot) == TRUE

dates <- subset(airquality, airquality$allfour == TRUE); dates
```

### d. Using the chi-square test

```{r}
tbl <- table(airquality$smoggy, airquality$windy)
Expected <- outer(rowSums(tbl), colSums(tbl))/sum(tbl); Expected
chisq.test(airquality$smoggy, airquality$windy)
```
The Ozone and Windiness factor does not seem to have any chance ($4.29\times 10^{-6}$) to arise by chance. Perhaps they are independent?

```{r}
tbl <- table(airquality$sunny, airquality$hot)
Expected <- outer(rowSums(tbl), colSums(tbl))/sum(tbl); Expected
chisq.test(airquality$sunny, airquality$hot)
```

This had one of the higher chances of arising by chance (2.408%).

































